Xolotl: An Intuitive and Approachable Neuron \& Network Simulator in MATLAB

### 2018-08-14 paper/paper.tex
inserted benchmark figure

Our goal in designing "xolotl" was to create an easy-to-use neuron and network simulator that was fast enough and accurate enough for routine use by computational neuroscientists. In this section we compare the speed of "xolotl" in simulating single and large numbers of Hodgkin-Huxley-like neuron models (as in Figure \ref{fig:figurehh}) and bursting neuron models based on the bursting neurons in the lobster STG (\cite{prinzSimilarNetworkActivity2004}). For each type of neuron model, we compared our software to  "NEURON" (\cite{hinesNEURONSimulationEnvironment1997}), a high-performance and powerful neuron simulator and "DynaSim" (\cite{sherfeyDynaSimMATLABToolbox2018}), a general-purpose simulator that can solve coupled differential equations numerically. All simulators were run on the same hardware using fixed time step solvers: "xolotl" used the Exponential Euler method  (\cite{dayanTheoreticalNeuroscience2001}), "NEURON" used the ALEC_FIX_ME and "DynaSim" used the ALEC_FIX_ME solvers. We measured the speed of each simulator by dividing the time simulated for by the time it took for the simulator to complete. For example, if a simulator could simulate 10 seconds of neuron data in 1 second, its speed would be 10X. We measured the speed of every simulator as a function of simulation time step, total length of simulation, and system size.

We also measured the

To assess speed and accuracy, "xolotl",  were compared in simulations over varied time resolution,  simulation time, and number of compartments (Figure \ref{fig:figurebenchmark}).

Single-compartment Hodgkin-Huxley-like models with three conductances and stomatogastric ganglion neuron models with eight conductances were generated using conductance dynamics from \cite{liuModelNeuronActivityDependent1998}) in the simulation environments. "xolotl" uses the exponential Euler method for integrating membrane potential (\cite{dayanTheoreticalNeuroscience2001}). "DynaSim" with "C"-compilation was implemented with a 2\textsuperscript{nd}-order Runge-Kutta integration scheme as recommended for high-performance in the documentation. "NEURON" used an implicit Euler regime (\cite{hinesNEURONSimulationEnvironment1997}).

To compare the integration methods, models were simulated for 30 s at varying time-resolution (Figure \ref{fig:figurebenchmark}A,E). The speed factor was defined as the ratio between time represented in the simulation and actual runtime (simulation-time). Therefore, the speed factor represents how many times faster the simulation is than a real-time observation. The error resulting from coarser time-resolution was computed as the distance between two LeMasson matrices (\cite{lemassonIntroductionEquationSolving2000}), one representing the canonical voltage trace with time step $\Delta t = 0.001$ ms and the other representing the voltage trace simulated at time steps up to $1$ ms (Figure \ref{fig:figurebenchmark}B,F).

To assess the performance of the simulators in absence of set-up overhead, models were simulated with a time-resolution of $0.1$ ms over increasing simulation time (Figure \ref{fig:figurebenchmark}C,G).

Many simulators perform well in simulations of many compartments (\cite{bretteSimulationNetworksSpiking2007, sherfeyDynaSimMATLABToolbox2018, vitayANNarchyCodeGeneration2015, delormeSpikeNETEventdrivenSimulation2003}). To assess how "xolotl" performs under these conditions, networks of up to 1,000 neuron models were simulated for 30 s at a time-resolution of 0.1 ms (Figure \ref{fig:figurebenchmark}D,H).

"xolotl" performed well in the benchmark simulations. Over increasing time step, "xolotl" was faster than "DynaSim" and "NEURON" with smaller error. During longer simulations "xolotl" and "DynaSim" increased in speed. This is likely due to the decreased impact of set-up overhead from "MATLAB", an interpreted language. For most simulation times, "xolotl" is roughly ten times faster than "DynaSim" and about twice as fast in the long-simulation time limit. "NEURON" performed consistently at high speed, likely due to a lack of programmatic overhead, though "DynaSim" and "xolotl" were significantly faster at large time steps and during long simulations.

@@

We set out to design a neuron and network simulator that could be useful in the classroom setting, especially for students of computational science, while also being powerful, fast, and extensible enough to be used for research. By using a novel architecture that permits the symbolic manipulation of "C++" objects in a comprehensible "MATLAB" interface, we showed how to create and simulate single-compartment models of Hogkin-Huxley like neurons (Figure \ref{fig:figurehh}), voltage clamp experiments and recover activation functions of single channel types (Figure \ref{fig:figureclamp}), a network of neurons with multiple synapse types (Figure \ref{fig:figurenetwork}), a multi-compartment neuron model where the voltage of every compartment is coupled, and a neuron model where homeostatic mechanisms control the maximal conductance of multiple channel types (Figure (\ref{fig:figuremechanism}).

\subsection{Design: components vs. equations in simulators }

- components from library vs. equations
- equations are transparent -- you know what you're solving. but clunky -- string parsing headaches, etc.
- components are easy to assemble, but it;s hard to know where they are, how they can be changed, etc.
- xolotl goes the component route, but makes it very easy to find components (you can click your way into the C++ files). components are stored as source files, so you can modify it and immediately see it reflected in your model. we didn't invent another synatax for these components -- these are just C++ files.

@@

\begin{quote}
	\textit{``About half the time spent on a typical simulation project involves creating and tuning the model. Thus, a good user interface may contribute more to the overall efficiency of a project than pure computation speed.''}
	\begin{flushright}
		\cite{deschutterConsumerGuideNeuronal1992}
	\end{flushright}
\end{quote}

"xolotl" is designed with ease-of-use in mind. This includes the time it takes to install, setup, and learn the software, the time to write and debug scripts, and the time to perform the simulations. An easy-to-use simulation environment must minimize time spent in all these domains, especially during human engagement with the software. Complicated software remains broadly inaccessible and time-consuming even to perform single-compartment simulations, though the actual simulation time may be very small.

The front-end interface to "xolotl" is built in "MATLAB" to facilitate interoperability with existing tools. "MATLAB" provides a comprehensive distribution from a single vendor with straightforward and powerful toolboxes for time series analysis, optimization, and parallel computing. For these reasons, "MATLAB" has remained a popular tool for data analysis and visualization in neuroscience. Since "xolotl" can be run directly from the "C++" executable, it would be possible to have implemented the interface in other popular scientific computing languages (\viz "Python" or "Julia"). Ultimately due to the popularity and ease of "MATLAB" for pedagogy and research, and its extensive support for vector and matrix mathematics, "xolotl" supports a "MATLAB" interface.

@@

>>>>

Our goal in designing "xolotl" was to create an easy-to-use neuron and network simulator that was fast enough and accurate enough for routine use by computational neuroscientists. In this section, we compare the speed of "xolotl" in simulating single and large numbers of Hodgkin-Huxley-like neuron models (as in Figure \ref{fig:figurehh}) and bursting neuron models based on the bursting neurons in the lobster STG (\cite{prinzSimilarNetworkActivity2004}). For each type of neuron model, we compared our software to  "NEURON" (\cite{hinesNEURONSimulationEnvironment1997}), a high-performance and powerful neuron simulator specialised in simulating neurons with complex morphologies and "DynaSim" (\cite{sherfeyDynaSimMATLABToolbox2018}), a general-purpose simulator that can solve coupled differential equations numerically. All simulators were run on the same hardware using fixed time step solvers: "xolotl" used the Exponential Euler method  (\cite{dayanTheoreticalNeuroscience2001}), "NEURON" used the  implicit Euler solver (\cite{hinesNEURONSimulationEnvironment1997}) and "DynaSim" used "C"-compiled 2\textsuperscript{nd}-order Runge-Kutta integration scheme as recommended for high-performance in the documentation. We measured the speed of each simulator by dividing the time simulated for by the time it took for the simulator to complete. For example, if a simulator could simulate 10 seconds of neuron data in 1 second, its speed would be 10X. We measured the speed of every simulator as a function of simulation time step, total length of simulation, and system size.

All three simulators were faster with larger time steps, since fewer iterations were needed (Figure \ref{fig:figurebenchmark}A-B), and were approximately linear in the region tested. Our simulator compared favourably to "NEURON" and "DynaSim" in this task. We also measured the quality of the simulated output by comparing it to the simulated output at the smallest time step. Simulation error was measured using the LeMasson cost (\cite{lemassonIntroductionEquationSolving2000}), and was comparable amongst the three simulators (Figure \ref{fig:figurebenchmark}C-D). Our simulator sets up and runs the simulation in "C++", and needs to transfer parameters and data to and from the underlying implementation. To measure the performance cost of this overhead, we repeated these simulations on all three simulators at a afixed time step of $0.1$ ms and varied the length of time simulated for. Speed increased with simulation duration up to a limit, and then saturated, indicating a fixed performance cost of the overhead (Figure \ref{fig:figurebenchmark}E-F, black lines). Simualtions using "DynaSim", which also used a similar architecture, showed a similar increase in speed with simulation length (Figure \ref{fig:figurebenchmark}E-F, red lines). In contrast, simulations using "NEURON" ran at a constant speed irrespective of simulation length, presumably due to differences in the underlying implementation (Figure \ref{fig:figurebenchmark}E-F, blue lines).

Many simulators have been designed to simualte large numbers of compartments, either as networks with many identical neurons or in a large mulit-compartmentel neuron model (\cite{bretteSimulationNetworksSpiking2007, sherfeyDynaSimMATLABToolbox2018, vitayANNarchyCodeGeneration2015, delormeSpikeNETEventdrivenSimulation2003}). While our software is not designed for this task {\em per se}, we measured its performance as a function of the number of compartments simulated. "xolotl" can quickly create a number of identical copies of a compartment using the "replicate" method:

\begin{lstlisting}[style=Matlab-editor]
x.replicate('compartment_to_replicate',n_copies); 
\end{lstlisting}

We used the "replicate" method to create simulations with varying numbers of neurons (either Hodgkin-Huxley-like or bursting neurons) and measured the speed of all three simulators as a function of system size. Plotting the speed normalised by the system size, we observed that the speed of integration of "xolotl" is linear with system size (Figure \ref{fig:figurebenchmark}G-H, black lines), for up to 1000 single-compartment neurons (up to 13,000 ODEs). Its performance compares favourably with that of "NEURON" and "DynaSim" as a function of system size (Figure \ref{fig:figurebenchmark}G-H).

@@

We set out to design a neuron and network simulator that could be useful in the classroom setting, especially for students of computational science, while also being powerful, fast, and extensible enough to be used for research. By using a novel architecture that permits the symbolic manipulation of "C++" objects in a intuitive "MATLAB" interface, we demonstrated some of the features of our simulator using simulations of single-compartment models of Hogkin-Huxley like neurons (Fig. \ref{fig:figurehh}); voltage clamp experiments to recover activation functions of single channel types (Fig. \ref{fig:figureclamp}); a neuron model where intracellular mechanisms can control the dynamics of Calcium and can regulate the maximal conductance of ion channel types in an activity-dependent manner (Fig. (\ref{fig:figuremechanism}); and a network of neurons with multiple synapse types (Fig. \ref{fig:figurenetwork}). We also illustrated how built-in features of our simulators make it easy to bookmark and jump between model configurations (Fig. \ref{fig:figuremechanism}, purple), and how parameters of the model can be changed using sliders and their effect can be viewed in real time (Fig. \ref{fig:figuremanipulate}).

\subsection{A focus on usability}

``About half the time spent on a typical simulation project involves creating and tuning the model. Thus, a good user interface may contribute more to the overall efficiency of a project than pure computation speed.'' \cite{deschutterConsumerGuideNeuronal1992} "xolotl" is designed primarily with ease-of-use in mind. This includes the time it takes to install, setup, and learn the software, the time to write and debug scripts, and the time to perform the simulations. An easy-to-use simulation environment must minimize time spent in all these domains, especially during human engagement with the software. Complicated software remains broadly inaccessible and time-consuming even to perform single-compartment simulations, though the actual simulation time may be very small.

We have focussed on making our software as easy to use as possible, without sacrificing performance or extensibility. For example, the software and all dependencies can be installed using a single-line installer script from within the "MATLAB" command line. The installation includes worked example scripts that demonstrate various features of the simulator, that can be run without any configuration.  We decided to built the the front-end interface to "xolotl" in "MATLAB" to facilitate interoperability with existing tools for time series analysis, optimization, and parallel computing. This allowed us to build rich tools for visualisation and to interact with the simulation.

@@

Over the years, several simulators have been developed to integrate systems of coupled differential equations,  those that model the spiking activity of neurons (\cite{bretteSimulationNetworksSpiking2007, sherfeyDynaSimMATLABToolbox2018, vitayANNarchyCodeGeneration2015, delormeSpikeNETEventdrivenSimulation2003, hinesNEURONSimulationEnvironment1997}). A critical design choice

@@

- components from library vs. equations
- equations are transparent -- you know what you're solving. but clunky -- string parsing headaches, etc.
- components are easy to assemble, but it;s hard to know where they are, how they can be changed, etc.
- xolotl goes the component route, but makes it very easy to find components (you can click your way into the C++ files). components are stored as source files, so you can modify it and immediately see it reflected in your model. we didn't invent another synatax for these components -- these are just C++ files.
- scaffolding
